---
title: "Learning"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages needed

```{r message=FALSE, warning=FALSE}
#install.packages(c("dplyr","haven","readr", "tidyr","ggplot2","texreg","stringr","stringi","WDI"))
library(haven)        # Import Data
library(readr)        # Import Data
library(dplyr)        # Data manipulation
library(ggplot2)      # Creating graphics
library(stringr)      # Deal with string data
library(kableExtra)   # Build complex tables and manipulate table styles
library(WDI)          # Get data from the World Bank's API
library(psych)        # We use the describe() function from this package for summary statistics.
library(texreg)       # Convert regression output to Table
library(stargazer)    # Convert regression output to Table
library(equatiomatic) # Provide an automatic way to show the equations of models fitted in R.
```

# Basic R

## R Data Type

What are the main types of data we will deal with?\
They can be:\
1. Numbers

```{r, results='hide'}
0
1.5
3/4
```

2.  Strings:

```{r, results='hide'}
"Hello world"
```

3.  Categories: [England, Scotland, Wales, Northern Ireland]

4.  Logical (MUST use uppercase):\

-   **TRUE**
-   **FALSE**\

Or we can use **T** and **F**.

## Creating objects

Data types can be contained in R objects.\
In RStudio we make vast use of the \<- operator to assign values to
objects. We can obtain the operator quickly by typing the "alt" + "-"
keys together.

```{r}
a <- 2
```

The above code assigns the value of '2' to an object called a.\
Notice that a is now in our *global environment* in the top-right part
of RStudio (unless you've changed your settings).

We can access the contents of a by running it as code:

```{r}
a
```

## Data Structures

There are many types of objects in R. Today we focus our attention on:\
1) Vectors\
2) Matrices (2-dimensional vectors)\
3) Data Frames\
4) Lists\

### Vectors

Vectors are one-dimensional objects that can have different lengths and
can contain different values, but all of the same nature (all numeric,
all characters, all logical...)

```{r}
my.first.vector <- c(1, 5, 2, 6, 8, 9, 12, 2.4, 0.3, -2.56, 3/4, 2+3)
# the c() function stands for "concatenate" and is used to join objects together
my.first.vector
```

How long is my vector?

```{r}
length(my.first.vector)
```

You can access the value of a specific element in your vector using
squared brackets:

```{r}
my.first.vector[8] # the value of the eighth element in our vector
my.first.vector[1] # the value of the first element in our vector
```

We can also store string elements in our vector:

```{r}
a.string.vector <- c("here", "are", "some", "elements")
a.string.vector
```

What data type is in the vector?

```{r}
class(my.first.vector)
class(a.string.vector)
```

paste() is a useful function to use with strings:

```{r}
c <- "hello"
d <- "world"
paste(c, d, sep = "")
c(c, d)
paste(c, d, sep = ",") # specifies the separator of the two objects!
```

paste0() is a similar function,which by default puts no separator:

```{r}
paste0(c, d)
```

Our vector can be a vector of factors. This is used to create
categorical variables.

### Matrix

A matrix is a two-dimensional vector. Besides being two-dimensional, the
same rules of vectors apply to matrix: a matrix can only contain data of
the same type.\
Suppose we want a 3x4 matrix with numbers from 1 to 12 in its cells:

```{r}
m <- matrix(data = 1:12,  # the data in our matrix. Pick all integers from 1 to 12 (included)
            nrow = 3,     # the number of rows 
            ncol = 4,     # the number of columns
            byrow = TRUE) # do we want data to be stored in cells by row or by column?
m
```

We can access an individual element of a matrix in R by specifying its
position.

```{r}
m[2,4] # the object in the 2nd row, 4th column
m[,1]  # all the first column
m[1,]  # all the first row
```

We can also give names to columns and rows.

```{r}
colnames(m) <- c("one", "two", "three", "four")
rownames(m) <- c("a", "b", "c")
m
```

I won't be showing you matrix algebra in action during this class
(you're unlikely to need it outside of introductory exercises and
advanced programming). But to summarise the available functions:

-   $+$ matrix addition
-   $*$ multiplication by a scalar
-   $%*%$ matrix multiplication
-   t( ) transpose
-   solve( ) inverse
-   diag( ) get the diagonal

### Data Frame

They are perhaps the most common object we will deal with in R. Most
databases are imported automatically as data frames. They have multiple
variables organized in columns, and each row represents one observation.

Before importing one, let's create one from scratch and see how it looks
like.

```{r}
datafr <- data.frame(
  var1 = c(rep("a", 3), rep("b", 4), rep("c", 2), "d"),
  var2 = rnorm(n = 10, mean = 0, sd = 1.3), # here we are drawing 10 random values from a 
  # normal distribution with mean = 0 and sd = 1.3
  var3 = c(1, -4, 2, 4.2, 5.3333, 1/9, 7.5, 0.000, 1-12, sqrt(2)) # notice that here we are
  # creating a vector where the elements are expressions! We also include the squared root of 2
)

datafr

```

Each variable within our data frame is a vector, and the rules of
vectors apply to them. All variables in a data frame must have the same
length. But we might have variables with missing values. In R these
missing values are called NA:

```{r}
df <- data.frame(var1 = c(rep("a", 3), rep("b", 4), rep("c", 2), "d"),
                 var2 = c(1,2,3,4,5,NA,7,8,9,10))
df # we have a missing value in var2

class(datafr)
```

We can assess information in one column of our data frame using **\$**
operator:

```{r}
datafr$var1 # this gives us precisely the object we want.
datafr$var3

datafr$var2
df$var2

class(datafr)
class(datafr$var1)
class(datafr$var2)
class(datafr$var3)
```

### List

Lists are the last object type that we briefly introduce today. Lists
are objects that include different other objects in R, of whatever type.
A list can even contain another list.

```{r}
ls <- list(df, datafr, datafr$var2)
ls

ls2 <- list(df, ls)
ls2
```

There are many functions that allow you to try to coerce the nature of
an object:\
- as.numeric() tries to coerce its argument into a numeric.\
- as.integer() tries to coerce its argument into an integer.\
- as.character() tries to coerce its argument into a character.\
- as.factor() tries to coerce its argument into a factor.\

And also its class! - as.matrix() - as.data.frame() - as.array() -
as.list()

# Import and Manage Datasets

## Import Datasets

Now, we will learn how to import different types of databases from
outside. First, let's clean our enviornment before dealing with
databases.

```{r}
rm(list = ls())
```

Or by clicking on session -\> clear workspace -\> yes.\

If you do it by code, make sure to keep that line **at the top** of your
script. It's really important that you start fresh each time you do work
in R, to make sure that your workflow (the order you do things) is
replicable. In other words, it helps make sure that your future self or
someone else can open up your script,run everything, and get the same
result every time.\

Then, we must set up our working directory, the folder on our computer
where we are going to work from, where we store our data and where
perhaps we might want to save our plots or the outputs from the
analysis.\

In R, we do it using the setwd() function:

```{r}
setwd("~/Desktop/ESS_3x_website")
```

Now import data from a csv file and save it in a named object.

The choice across various functions of importing data depends on the
data type, e.g. **.dta**, **.csv**, or **.xlsx**.

```{r}
ANES <- read_dta("anesByState.dta")
head(ANES)
```

```{r}
csvfile <- "https://raw.githubusercontent.com/lorenzo-crippa/Intro_to_R/master/2021/baseball.csv"
baseball <- read.csv(csvfile)
head(baseball)
```

Yes! We could import a data file from a URL:

```{r}
url <- "http://www.principlesofeconometrics.com/stata/broiler.dta"
data.df <- read_dta(url)
head(data.df)
```

In addition, although we won't get into the details of this method, some
packages allow users to search and download data, e.g. **WDI**, **V-Dem,
democracyData**. Here is an example of importing World Bank Data
automatically.

```{r}
wdi_data <- WDI(country = "all", 
               indicator = c(population = "SP.POP.TOTL",
                             gdp_percapita = "NY.GDP.PCAP.KD"), 
               start = 2000, end = 2020, extra = TRUE)
head(wdi_data)
```

Here is the common data types, and the corresponding functions and
packages.

| Data Type | Function   | Package |
|-----------|------------|---------|
| **.dta**  | read_dta   | haven   |
|           | read.dta   | foreign |
| **.csv**  | read_csv   | readr   |
|           | read.csv   |         |
| **.xls**  | read_excel | readxl  |
| **.xlsx** | read_excek | readxl  |
| **.txt**  | read.delim |         |

## Manage datasets

Database management like dropping observations and variables,sorting the
data and the other, can be done using the basic R language.\

However, a suite of packages has also been developed to write code in R
in much more natural syntax. The suite is called tidyverse. We can
install it by running:\

In what follows, you'll see code for managing datasets in base R,and
next you'll see the (more intuitive) tidyverse version of the same
operations. It's important to know both.

Suppose now we want to drop from our baseball data all observations that
refer to the CLE team (no offense to its fans). Doing it in base R would
be:

```{r}
baseball.no.CLE <- baseball[baseball$team != "CLE", ]
```

Suppose now we want to get rid of all observations that are not for New
York Yankees (NYY) and also we want to drop the column "name". Yet, we
don't want to overwrite the "baseball" object because we might need the
information we're dropping later. We can do the following:

```{r}
baseball.2 <- baseball[baseball$team == "NYY", 2:4]
```

These operations are made smoother by tidverse Let's do something
similar on ANES. Let's remove all observations from the year 2000, then
remove the variable turnout:

```{r}
ANES <- filter(ANES, # what is your data frame? 
               year != 2000) # the condition you impose
ANES <- select(ANES, 
               -turnout) # keep everything BUT the variable named "turnout"

```

Now suppose we want to sort observations in our ANES dataset based on
state, then year.We can do it by:

```{r}
ANES[order(ANES[,2], ANES[,1]),]
```

By doing this we have taken ANES and we have asked R to reorder its rows
first based on the second column (state), then by its first column
(year). Notice that this time we have not saved the data frame obtained
into an object!

Another useful function is unique(). It tells us what values a vector
takes:

```{r}
unique(ANES$year) # all possible years in the ANES dataset
# we might want to have it sorted:
sort(unique(ANES$year)) # years from 1952 to 2008 by 2-year distance. 

```

We can also add variables to our data frames. What if we wanted to add
the squared valueof "poor" in the ANES data frame? In base R this is
done by doing:

```{r}
ANES$new.var <- ANES$poor^2

# we can also do it in tidyverse by doing:
ANES <- mutate(ANES,
               new.var2 = poor*2)
```

ifelse() is a function that is very useful to know. If its "test"
argument (a condition) is verified, it does what the "yes" argument
says. Otherwise, it executes the "no" argument. For instance, in ANES we
might want to have a variable that takes value 0, if the year is before
2000, and 1 otherwise. We can do it like this:

```{r}
ANES$indicator <- ifelse(test = ANES$year >= 2000, yes = 1, no = 0)
unique(ANES$indicator)
```

# Get Summary Stats and Plots

## Summary Statistics

When we first obtain a data object in R, there are several things that
we could do:

```{r}
names(ANES)   # obtain all the variable names of a dataset
head(ANES)    
glimpse(ANES) # get a glimpse of your data.
```

If we want to get some description of our data.

```{r}
summary(ANES$poor)
```

We can also ask for specific information:

```{r}
mean(ANES$voteDem) 
sd(ANES$poor)
median(ANES$dem)
min(ANES$year)
max(ANES$voteDem)
quantile(ANES$poor, 0.3) # the observation that leaves 30% of the observations to its left
quantile(ANES$poor, 0.5) # this will be the median of course!
median(ANES$poor)
var(ANES$dem) # the variance
table(ANES$year) # how often does each occurrence appear?
range(ANES$year)
```

## Basic plots

We now want to obtain some plots about our data. We will see how to:

1)  obtain boxplots

2)  obtain histograms

3)  obtain density plots

4)  obtain twoway plots

### Boxplots

To create a boxplot:

```{r}
boxplot(ANES$white)

# we can then intervene on options and modify its appearance:
boxplot(ANES$white, frame = F, ylab = "white", main = "A boxplot")
?boxplot
```

We could also:

```{r}
# this would be the tidyverse way of doing it:
ggplot(ANES, aes(y = white)) +
         geom_boxplot() +
         theme_minimal()
```

### Histograms

Plotting a histogram is also easy and similar to create a boxplot:

```{r}
hist(ANES$FTM, main = "Histogram of FTM", xlab = "FTM")

# and we can also intervene in the plot with options:
hist(ANES$FTM, main = "Histogram of FTM", xlab = "FTM", probability = TRUE) # to have probabilities
```

Again, in tidyverse:

```{r}
ggplot(ANES, aes(x = FTM)) +
  geom_histogram(binwidth = 5) +
  theme_minimal()
```

### Density Plots

We can also obtain kernel densities of our data:

```{r}
density(ANES$dem)
plot(density(ANES$dem))
```

### Twoway Plots

Suppose now we want to explore two-way relations in our data. For
instance the two-way relation between height and weight of baseball
players:

```{r}
plot(x = baseball$heightinches, y = baseball$weightpounds,
     frame.plot = FALSE, xlab = "height", ylab = "weight",
     pch = 20) # pch is an option that selects the shape of dots we want 

# what if we want to add lines on this plot? We use the abline() function
# Maybe we want to plot a line corresponding to the mean height 
# lwd makes it thicker
abline(v = mean(baseball$heightinches, na.rm = TRUE), col = "red", lwd = 5) 
```

Again, tidyverse makes things much easier. Let's re-do all the steps we
did before. First, we created a scatter plot. This time, let's save the
plot in an object:

```{r}
my.plot <- ggplot(baseball, aes(x = heightinches, y = weightpounds)) +
  geom_point() +
  theme_minimal()
my.plot
```

tidyverse makes it also simpler to save tidyverse plots:

```{r}
ggsave(my.plot, filename = "my_first_ggplot.pdf")
```

# Run Basic Statistical Analyses

## T-Tests

Suppose we would like to run a test to see if the population mean of a
random variable is distinguishable from 0 based on the data:

```{r}
t.test(ANES$poor) # by defaul it's a two-tailed t-test with null-value = 0
```

Or you can also test if the true value is equal to other values:

```{r}
t.test(ANES$FTM, mu = 60)

# ...and you can ask the test be one-sided:
t.test(ANES$FTM, mu = 60, alternative = "less")
t.test(ANES$FTM, mu = 60, alternative = "greater")
```

We could also run difference-in-means tests. Suppose you want to test if
the mean height of players in the NYY team is different from the mean
height of players in the MIN team:

```{r}
t.test(x = baseball$heightinches[baseball$team == "NYY"],
       y = baseball$heightinches[baseball$team == "MIN"])
```

**YES, THEY ARE**. Players in the Yankees are taller on average.

## Statistical models

### Linear Models

Now, we want to estimate a linear model using OLS. To do that, we first
import a dataset from an article by David Card (1993) on NBER:

```{r}
card <- read_dta("Card_data.dta")
head(card)
```

Okay, let's run a linear regression.

```{r}
model1 <- lm(wage ~ exper, # your formula
             data = card) # where are the data stored
summary(model1)
```

We probably would like to include some control variables in this model
specifications.

```{r}
model2 <- lm(wage ~ exper + motheduc + fatheduc,
             data = card)
summary(model2)
```

From time to time, we need to transform some variables:

```{r}
model3 <- lm(log(wage) ~ exper + motheduc + fatheduc,
             data = card)
summary(model3)
```

And we could use extract_eq() to obtain the equation of the model
output:

```{r}
extract_eq(model3,wrap = TRUE, use_coefs = TRUE)
```

There are some very good packages to produce neat regression tables out
there to present our results. The choice between them depends mostly on
the type of models that they can support. Two equally good choices are
**stargazer** and **texreg**.

We'll use stargazer today:

```{r}
# let's see how stargazer works
stargazer(model1, model2, model3, type = "text",
          keep.stat = c("n", "rsq", "adj.rsq", "f"), df = FALSE)
?stargazer
```

```{r}
# we can also extract statistics and information from our model objects:
model3$coefficients
# model3$residuals
fit <- model3$fitted.values

# and more:
summary(model3)$coefficients[,1] # these are the coefficients (the betas)
summary(model3)$coefficients[,2] # these are the standard errors of our coefficients
summary(model3)$coefficients[,3] # these are the t-statistics
summary(model3)$coefficients[,4] # these are the p-values
# of course we could save this information in R objects
```

### Generalized Linear Models

Most statistical models in R have a syntax very similar to the one
adopted by lm().Covering these models is not part of today's class, but
we might want to show that other models, such as those in the family of
the generalized linear model (GLM),work very similarly. Suppose we want
to explain if a person went to college based on the years of education
of the father and mother. A logit or probit model would be appropriate:

```{r}
mod.glm <- glm(enroll ~ fatheduc + motheduc +
                 black + smsa + south + IQ,
               data = card,
               family = binomial(link = "logit"))
summary(mod.glm)
screenreg(mod.glm)
```

# Additional Sources

## Some additional packages

| Type            | Package Name         | Aim                                                                      |
|-------------------|-------------------|----------------------------------|
| Data import     | readr                | Import files from various formats                                        |
|                 | foreign              |                                                                          |
| Data Cleaning   | reshape              | Turning long datasets into wide and vice versa                           |
| Plots           | ggplot2              | Plots based on grammar of graphics                                       |
|                 | ggpubr               | ggplot2-based publication-ready plots                                    |
| Standard errors | sandwich + lmtest    | Robust and cluster + test coeffcients                                    |
| Modelling       | plm                  | Panel data models                                                        |
|                 | margins or ggeffects | Marginal effects from models                                             |
|                 | marginaleffects      | Marginal effects from models                                             |
|                 | AER                  | Various (Instrumental variable, tobit models)                            |
|                 | mlogit or mnlogit    | Multinominal logit models                                                |
|                 | lme4                 | Multilevel models                                                        |
|                 | rdd                  | Regression discontinuity design                                          |
|                 | zoo or forecast      | Time series tools                                                        |
| Latent Variable | psych                | Exploratory Factor Analysis                                              |
|                 | lavaan               | CFA/SEM                                                                  |
| Webscraping     | rvest                | Webscraping                                                              |
|                 | httr                 | Webscraping, APIs                                                        |
|                 | RSelenium            | Scraping dynamic websites                                                |
| Text Data       | quanteda             | Text data mining and wrangling                                           |
| Twitter Data    | rtweet               | Mining Tweets                                                            |
| Spatial Data    | tmap                 | Drawing thematic maps                                                    |
|                 | sf                   | Spatial Manipulaytion                                                    |
| Color Palette   | MetBrewer            | Palettes inspired by works at the Metropolitan Museum of Art in New York |

## User-written functions

User-written **functions** allow you to create a function which runs a
series of commands, it can include and input and output variable. You
must also name the function yourself. Once the function has been run, it
is saved in your work environment by R and you can run data through it
by calling it as you would do with any other function in R (I.e.
myfunction(data)).

The general structure looks like:

## Loop

A loop is a means to repeat code under defined conditions. This is a
great way for reducing the amount of code you may need to write. For
example, you could use the loop to iterate a series of commands through
a number of different positions (i.e. rows) in a data frame.

The most basic means of looping commands in R is known as the for loop.
To implement it, you run the `for()` function and within the parameters,
you define the sequence through which the loop will input data. Usually,
an argument is assigned (i) and given a range of numerical positions
(i.e. `in 2:5` which runs from position 2 to position 5). You next enter
the commands which the loop will run within curly brackets following the
parameters. The defined argument (i) is used to determine the input data
for each iteration. This has been demonstrated in the example below
which loops the code for four columns of our data frame.

## If/else statements

There are various commands we can run within for loops to edit their
functionality. If-else statements allow us to change the outcome of the
loop depending on a particular statement. For instance, if a value is
below a certain threshold, you can tell the loop to round it up,
otherwise (else) do nothing.

# Appendix: R Script

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
